---
title: "Hotel Booking Analysis"
author: Charlie Logan, Jonathan Demeter, Ana Mohapatra, Nikita Mehendale, Srishti Senthil
date: February 19, 2021
output: 
html_document:
toc: true
toc_depth: 2
toc_float: true
number_sections: true
fig_caption: true
--

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Table of Contents

## Introduction

```{r echo=FALSE}
hotels_bookings <- read.csv("hotel_bookings.csv")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(ggplot2)
theme_set(theme_minimal())
library(dplyr)  #needed for dataset row/column functions
library(viridis)
library(patchwork)
library(hrbrthemes)
library(methods)

hotels_bookings$arrival_date_year<-as.factor(hotels_bookings$arrival_date_year)
hotels_bookings$arrival_date_month<-as.factor(hotels_bookings$arrival_date_month)
hotels_bookingsarrival_date_day_of_month<-as.factor(hotels_bookings$arrival_date_day_of_month)
hotels_bookings$is_repeated_guest<-as.factor(hotels_bookings$is_repeated_guest)
```
## Data exploration

What is the average number of days in waiting list?
```{r echo=FALSE}
mean(hotels_bookings$days_in_waiting_list)
```
## Arrivals

### Arrivals by Month

In the following chunk we generate a frequency distribution to display the proportion of total bookings per month based on arrival date.

```{r arrive, echo=FALSE}
#Create a frequency table with 12 rows: month names; # arrivals. Then plot.
barplot(sort(prop.table(table(hotels_bookings$arrival_date_month))), main="Proportion of Arrivals by Month", ylab="Percentage of Total Arrivals",las=2, cex.names=0.7, col = "darkolivegreen", border = "darkslategray",)

```

From the above plot, it is clear that guests book the greatest number of stays arriving during the peak summer months of July and August.
Guests book the fewest number of stays to arrive during the winter months of January and December. Hotels may consider events or promotions to promote bookings during these slower months. Hotels could also lower their prices or cut down on operational costs during those months to maximize profit.

### Arrivals by Week
In the following chunk we plot how many bookings arrive each week as a bar chart. 
```{r arr_by_wk, echo=FALSE}
#As line graph:
weekly_arrivals <- as.data.frame(table(hotels_bookings$arrival_date_week_number, dnn = list("Week Number")), responseName = "Frequency of Arrivals")

lg <- ggplot(data=weekly_arrivals, aes(x=Week.Number, y=Frequency.of.Arrivals))+
  geom_point(shape=18, color="blue") 



bg <- ggplot(data.frame(hotels_bookings$arrival_date_week_number), aes(x=hotels_bookings$arrival_date_week_number, fill=factor(hotels_bookings$arrival_date_month, levels = month.name)) ) +
         geom_bar() +
          theme_ipsum() +
              labs(title="Arrivals Frequency by Week", 
              x="Week Number", y = "Arrivals (in Number of Bookings)")+ 
              guides(fill=guide_legend(title="Arrival Month"))
```

This barchart combines the data we have already seen to follow arrivals frequency over time. The arrivals by week gives a closer look at popular weeks for people to book reservations. The hotel may investigate these patterns to prepare accordingly. This could be purchasing supplies according to expected volume.

### Peak, Normal, and Slow Weeks
In the following chunk "season" we assign "Peak", "Normal", and "Slow" levels to a new column in the weekly_arrivals data based on the frequency of arrivals in that week number.
```{r season, echo=FALSE}

#define categorical variable to say whether week number falls during slow, normal, or peak season.
summary(weekly_arrivals$Frequency.of.Arrivals) # We will apply a logical condition to this column's vales to assign season

weekly_arrivals$season <- NA
for( i in rownames(weekly_arrivals) ) {
   if(weekly_arrivals[i, "Frequency.of.Arrivals"] >= 2500){
     weekly_arrivals[i,"season"] <- "Peak"
   } else if (weekly_arrivals[i,"Frequency.of.Arrivals"] > 1500 & weekly_arrivals[i,"Frequency.of.Arrivals"] < 2500){
    weekly_arrivals[i,"season"]  <- "Normal" 
   } else {
    weekly_arrivals[i,"season"]  <- "Slow"
  }
}
weekly_arrivals$season <- as.factor(weekly_arrivals$season)
nlevels(weekly_arrivals$season)  #verify that season has 3 levels

```
We have now grouped the by-week arrivals data into "season" labels.

### Regression Models for Arrivals Frequency  

```{r lin, echo=FALSE}
lm1 <- lm( Frequency.of.Arrivals ~ season,data=weekly_arrivals)
summary(lm1)
```
We can interpret the above regression model as follows:
Frequency of arrivals is explained by season per the linear function: Frequency.of.Arrivals = 2074 + 733*(peak_season,1,0) - 847.5(slow_season,1,0)

If the week changes from a normal season to a peak season, the hotels can estimate an aggregate 733 additional arrivals that week. If the week changes from a normal season to a slow season, the hotels can estimate an aggregate 848 fewer arrivals that week. This may be useful for quick decision-making in purchasing or event management by hotel management.
The adjusted r-squared value indicates that season explains 78.6% of the variance in frequency of arrivals, which is reasonably significant.


Arrivals is an interesting distribution which he have now visualized in several different plots and models. We will seek to compare this data to price information in the following section.

## Pricing
The average daily rate across all reservations in both hotels is as follows:
```{r adr_city_res, echo=FALSE}
adr <- tapply(hotels_bookings$adr, hotels_bookings$hotel, mean, na.rm = TRUE)
adr
```
This table shows us that on average, customers pay more per day at City Hotel rather than Resort Hotel.

Now let's look at how different factors effect the total price that customers have to pay for their stay.
Does the distribution channel have an effect on how much a customer pays?
```{r, echo=FALSE}
hotels_bookings$priceperstay <- hotels_bookings$adr *(hotels_bookings$stays_in_weekend_nights + hotels_bookings$stays_in_week_nights)
dist <- tapply(hotels_bookings$priceperstay, hotels_bookings$distribution_channel, mean, na.rm = TRUE)
sort(dist)
```
It appears corporate distribution channel pays the least while the direct distribution channel leads to the highest price.

Does paying a deposit have an effect on how much a customer pays?
```{r, echo=FALSE}
depo <- tapply(hotels_bookings$priceperstay, hotels_bookings$deposit_type, mean, na.rm = TRUE)
barplot(sort(depo), col=rgb(0, .1, .7), main="Average Price paid Vs. Deposit Type", xlab="Deposit Type", ylab="Price Paid")
```

We can see that the highest price is charged when there is no deposit and the lowest price is paid when the deposit was Non Refundable.

Now we wish to explore whether higher prices are charged in peak seasons.
In the following chunk we will first look at the ADR distribution to check for outliers and prepare the data for analysis.
```{r adr, echo=FALSE}
adr_week <- ggplot(data=hotels_bookings, aes(x=arrival_date_week_number, y=adr))
adr_week + geom_point() +
              labs(title="Average Daily Rate by Week", 
              x="Week Number", y = "ADR (in USD)")    #Note the one crazy ADR outlier around week 13 or so! Got charged $$$

# boxplot(hotels_bookings$adr, horizontal = T, col = "darkseagreen1", xlab = "Average Daily Rate (in USD)")   # Can see that outlier where ADR >$5K
hotels_bookings$adr_adj <- ifelse(hotels_bookings$adr<1000, hotels_bookings$adr, NA) # Set outlier to NA
boxplot(hotels_bookings$adr_adj, horizontal = T, col = "darkseagreen2", xlab = "Average Daily Rate (in USD)")  # Verify that ADR outlier has been eliminated


```
  

In the next chunk, we will visualize the movement in this distribution by plotting vertical boxplots of the average daily rate against week number.
```{r adr_boxplots, echo=FALSE}
boxplot(adr_adj ~ arrival_date_week_number, data=hotels_bookings, main="Average Daily Rate throughout the Year", xlab="Week Index", ylab="Average Daily Rate in USD (ADR)" )
```

Already, the points appear to be higher in mid-year (peak summer) and at year-end (around Christmas through New Year's). It is clear that the thick middle lines (denoting the median) hike upward to higher ADR around Weeks 30-34 and surely for Week 53, confirming our hypothesis. In general, all boxplots are skewed slightly right, since each week, a few bookings were significantly higher-priced than the critical mass.

Now the ADR data is ready to be compared with weekly arrivals.

```{r adr_and_arrival, echo=FALSE, message=FALSE, warning=FALSE}

#compute average adr for each week, store this as column in Weekly Arrivals
weekly_arrivals$adr <- NA  # fix dummy value to populate new column
  for (wk in weekly_arrivals$Week.Number) {
  this_week_in_h <- subset(hotels_bookings, hotels_bookings$arrival_date_week_number == wk )
  # print(summary(this_week_in_hotels_bookingsadr_adj)) # Verify this subset of h is indeed returning a different distribution in each week
  weekly_arrivals[wk,"adr"] <- mean(this_week_in_h$adr_adj)   #mean of ADRs in subset of h corresponding to each week
  }

sum(!complete.cases(weekly_arrivals$adr[-1]))   # Note that 0 still contain NA after the for-loop calculates each week's mean ADR
summary(weekly_arrivals$adr) # The spread of weekly average ADRs

coeff <- 20
adr_arr <- ggplot(weekly_arrivals, aes(x=Week.Number)) + 
  geom_line(aes(y = adr), color = "darkred") +    # Mean of ADRs for bookings with arrival dat in that week
  geom_line(aes(y = Frequency.of.Arrivals / coeff), color="steelblue", linetype="twodash") + # number of arrivals in that week
  xlab('Week Index') +

  scale_y_continuous(                       #Custom y-scales
    name = "Mean of ADRs",
    sec.axis = sec_axis( trans=~.*coeff, name="Frequency of Arrivals")
  ) +
  theme_ipsum()+
    theme(
    axis.title.y = element_text(color = "darkred", size=13),
    axis.title.y.right = element_text(color = "steelblue", size=13)
  ) +
  ggtitle("ADR and Arrivals Frequency by Week")

adr_arr
```

The double line chart demonstrates a high correlation between ADR and arrivals frequency through the course of a year. Rates are hiked during busier weeks, and rates are lowered during slower weeks. The hotels are able to capitalize on an opportunity to maximize revenue during these high-demand weeks.

To make some quantitative statements regarding the relationship between these two response variables, we can compare ADR across peak, normal, and slow weeks (in terms of arrivals frequency) with a table as follows:

```{r adr_vs_peak, echo=FALSE}

# Take the weekly_arrivals data, then group_by season, and summarize by mean adr.

weekly_arrivals %>%
  group_by(season) %>%
  summarize(count_of_wks = n(),
            mean_adr = mean(adr, na.rm=T))

```

On average across both hotels, \$87 is charged nightly during normal season, \$117 is charged nightly during peak season, and $70 is charged nightly during the slow season.

## Companies and Agents

Which companies have the highest mean price per stay for City Hotel?
```{r echo=FALSE}
h_city <- hotels_bookings[hotels_bookings$hotel == "City Hotel",]
cityhotelprice <- tapply(h_city$priceperstay, h_city$company, mean, na.rm = TRUE)
barplot(sort(cityhotelprice, decreasing = TRUE)[1:20], col=rgb(0, .1, .6))

```

Which companies have the highest mean price per stay for Resort Hotel?

```{r echo=FALSE}
h_resort <- hotels_bookings[hotels_bookings$hotel == "Resort Hotel", ]
resorthotelprice <- tapply(h_resort$priceperstay, h_resort$company, mean, na.rm = TRUE)
barplot(sort(resorthotelprice, decreasing = TRUE)[1:20], col=rgb(0, .1, .6))
```
#So? elaborate

Which companies have the highest number of bookings for City Hotel?

```{r echo=FALSE}
barplot(sort(table(h_city$company), decreasing = TRUE)[2:21], col=rgb(0, .1, .6))
```

Which companies have the highest number of bookings for Resort Hotel?

```{r echo=FALSE}
barplot(sort(table(h_resort$company), decreasing = TRUE)[2:21], col=rgb(0, .1, .6))
```

Using the above data, City and Resort hotel can decide which companies to deepen their ties with based on the mean revenue and number of bookings they get from each company. 
For example, City Hotel can choose to provide discounts to the top 10 companies' employees in order to encourage loyalty (and more bookings) from that company. 

Which agents have the highest mean price per stay for City Hotel?

```{r echo=FALSE}
cityhotelagentprice <- tapply(h_city$priceperstay, h_city$agent, mean, na.rm = TRUE)
barplot(sort(cityhotelagentprice, decreasing = TRUE)[1:20], col=rgb(0, .1, .6))
```

Which agents have the highest mean price per stay for Resort Hotel?
```{r echo=FALSE}
resorthotelagentprice <- tapply(h_resort$priceperstay, h_resort$agent, mean, na.rm = TRUE)
barplot(sort(resorthotelagentprice, decreasing = TRUE)[1:20], col=rgb(0, .1, .6))
```

Which agents have the highest number of bookings for City Hotel?

```{r echo=FALSE}
barplot(sort(table(h_city$agent), decreasing = TRUE)[2:21], col=rgb(0, .1, .6))
```

Which agents have the highest number of bookings for Resort Hotel?

```{r echo=FALSE}
barplot(sort(table(h_resort$agent), decreasing = TRUE)[2:21], col=rgb(0, .1, .6))
```

Using the above data, the City and Resort hotel can decide which agents to deepen their ties with based on the mean revenue and number of bookings they get from each agent.
For example, Resort can choose to provide discounts to the top 10 agents in terms of bookings in order to encourage loyalty from those agents. 


## Cancellations
### Cancellation Rate for each Hotel
Of the two hotels in our dataset, City Hotel and Resort Hotel, which has a higher average cancellation rate?
```{r echo=FALSE}
hotels_bookings$hotel <- as.factor(hotels_bookings$hotel)
canceledbyhotel <- round(tapply(hotels_bookings$is_canceled, hotels_bookings$hotel, mean, na.rm=TRUE),digits = 4)*100
canceledbyhotel
```
City hotel has a higher cancellation rate, about 42% of reservations are canceled, while only about 28% of bookings are cancelled at Resort Hotel. This is valuable base statistic to know how many people that booked a reservatin will actually end up staying at each hotel. This is valuable because if each hotel knows about how many people will be staying during a given period they can plan their supplies accordingly to maximize profit.

### Cancellations by Nationality
We can compare the rate of cancellations by people from each country on a graph
```{r echo=FALSE}
hotels_bookings %>% group_by(country) %>% summarise(num_cancel = order(mean(is_canceled))) %>% head(10) %>% ggplot(aes(x = country, y = num_cancel, fill= num_cancel))+ geom_col()

```

People from Portugal have

### Cancellations by month

```{r echo=FALSE}
hotels_bookings %>% group_by(arrival_date_month) %>% summarise(num_cancel = sum(is_canceled)) %>% arrange(desc(num_cancel)) %>% head(10) %>% print %>% ggplot(aes(x = arrival_date_month, y = num_cancel, fill= num_cancel))+ geom_col()

hotels_bookings %>% group_by(arrival_date_month) %>% summarise(mean_cancel = mean(is_canceled)) %>% arrange(desc(mean_cancel)) %>% head(10) %>% print %>% ggplot(aes(x = arrival_date_month, y = mean_cancel, fill= mean_cancel))+ geom_col()

barplot(sort(table(hotels_bookings$arrival_date_month)), las=2, col=rgb(0, .1, .6), xlab="Arrival Date", ylab="Number of Bookings", main="Number of Bookings meant to Arrive by Month")
```
We can see that August and July have the most amount of cancellations based on the graph below. But the months of June and April have the highest rates of cancellation. This difference can be explained because August and July have the most amount of reservations, so it would make sense they also have a high cancellation count. J

The market segment with the highest cancellation count is the online travel agency market segment. The highest cancellation rate however is the Groups segment (besides Undefined segment because there are only two instances of that which isnt substantial enough to make conclusions on)
```{r echo=FALSE}
hotels_bookings %>% group_by(market_segment) %>% summarise(num_cancel = sum(is_canceled)) %>% arrange(desc(num_cancel)) %>% print %>% ggplot(aes(x = market_segment, y = num_cancel, fill= num_cancel))+ geom_col()

hotels_bookings %>% group_by(market_segment) %>% summarise(num_cancel = mean(is_canceled)) %>% arrange(num_cancel) %>% ggplot(aes(x = market_segment, y = num_cancel, fill= num_cancel))+ geom_col()
```

People book their hotel reservations primarily by online travel agencies
```{r echo=FALSE}
hotels_bookings %>% group_by(market_segment) %>% count %>% print %>%
          ggplot(aes(x = market_segment, y = n, fill= n))+ geom_col()
```

Is staying a fewer numbers of days more likely to be canceled then staying a week? First we explore how cancellation is affected by duration of stay. By looking at the first graph it isnt so clear whether a longer stay is less likely to be canceled because of the high cancellation for 12 day bookings. When we break up the data into 7 days stays or not, we can see there is a slightly higher probability of a stay less than a week being cancelled.
```{r echo=FALSE}
hotels_bookings %>% select(is_canceled, stays_in_weekend_nights, stays_in_week_nights) %>% 
                    mutate(days_stayed = stays_in_weekend_nights + stays_in_week_nights) %>%
                    group_by(days_stayed) %>% summarize(num_cancel = mean(is_canceled)) %>% print %>%
                    head(15) %>% ggplot(aes(x = days_stayed, y = num_cancel, fill = num_cancel)) + geom_col()


hotels_bookings %>% select(is_canceled, stays_in_weekend_nights, stays_in_week_nights) %>% 
                    mutate(days_stayed = stays_in_weekend_nights + stays_in_week_nights,
                           week_stay = ifelse(days_stayed >= 7, 1, 0)) %>%
                    group_by(week_stay) %>% summarize(num_cancel = mean(is_canceled))
```

Are guests staying only during the week more likely to cancel than guest staying only during the weekend? After filtering out the data to discern whether the reservation took place only during the week and not on the weekend, we an see that weekend only bookings were canceled at a significantly higher rate. 38% vs 27%, weekend vs weekday only respectively

```{r echo=FALSE}
hotels_bookings %>% select(is_canceled, stays_in_weekend_nights, stays_in_week_nights, reservation_status) %>%
                    filter(stays_in_week_nights != 0 | stays_in_weekend_nights != 0 ) %>% 
                    filter(stays_in_week_nights == 0 | stays_in_weekend_nights == 0) %>% 
                    mutate(weekDayOnly = ifelse(stays_in_weekend_nights == 0, 1, 0)) %>% group_by(weekDayOnly) %>%  
										summarize(mean_cancel = mean(is_canceled))
                    
```

Do deposits make it less likely for someone to cancel? Strangely, 99% of Non Refundable deposits ended up being canceled. This seems counter intuitive and a ridiculously high cancellation rate regardless of the context, especially because the count of Non REfundable deposits is ~14k which is very high. This leads me to believe this data is corrupted 

```{r echo=FALSE}
# QUESTION: All of (~99%) the non-refundable depoists cancelled their reservations (seems wrong)
#hotels_bookings %>% select(is_canceled, deposit_type) %>% filter(deposit_type == "Non Refund")
hotels_bookings %>% select(is_canceled, deposit_type) %>% group_by(deposit_type) %>% summarise(num_cancel = sum(is_canceled))


hotels_bookings %>% select(is_canceled, deposit_type) %>% group_by(deposit_type) %>% count

hotels_bookings %>% select(is_canceled, deposit_type) %>% group_by(deposit_type) %>% summarise(num_cancel = mean(is_canceled))
```

Are groups booking with children more or less likely to cancel their reservation? The table below indicates that groups with children and without children cancel reservations at about the same rate, at about 37% for parties without children and about 34% for those with children. 
```{r echo=FALSE}
hotels_bookings$has_children <- hotels_bookings$children > 0 | hotels_bookings$babies > 0
canceledbychilden <- aggregate(is_canceled ~ has_children, data = hotels_bookings, mean)
canceledbychilden
```

Traveling with children indicates a group is slightly less likely to cancel their reservation, but does that pattern change if the guest is traveling with a baby and no older children? The table below shows the difference in cancelation rates between parties with more than one baby and parties with either no babies or other children. 
```{r echo=FALSE}
hotels_bookings$has_baby <- hotels_bookings$babies > 0 & hotels_bookings$children == 0
(canceledbybaby<- aggregate(is_canceled ~ has_baby, data = hotels_bookings, mean))
```

Those with one or more babies in their party but no children are much less likely to cancel a reservation, at a cancelation rate of about 17% as compared to a party with just adults or adults and children canceling at a rate of about 37%. Young families with babies seem to be more reliable in their bookings, which is valuable information for a hotel to have when planning availabity and projecting which reservations are going to fall through. 


In addition to information about who is in a given party, hotels save information about guests previous cancelation behavior. Is a guest that has cancelled a reservation in the past more likely to cancel again? The table velow indicates that a guest who canceled a reservation previously is highly likely to cancel again, showing a cancelation rate of about 91% for prior cancellers and a below average rate of about 34% for guests who have no record of previous cancelation. 

```{r echo=FALSE}
hotels_bookings$has_previous_cancellation <- hotels_bookings$previous_cancellation != 0
(repeat_cancel<- aggregate(is_canceled ~ has_previous_cancellation, data = hotels_bookings, mean))
```

The barplot pictured below shows the cancelation rate of these two groups. If hotels are concerned about cancelations and would like to improve their ability to predict which guests will not end up staying at the hotel, ensuring guests who have canceled in the past do not repeat would be a good place to start. 

```{r echo=FALSE}
p <- ggplot(data=hotels_bookings, aes(x=has_previous_cancellation, y = is_canceled, fill=has_previous_cancellation)) + stat_summary(fun = "mean", geom="bar")
p + ggtitle("Plot Reservation Cancellation Rate by Previous Cancellation") +
  xlab("Has Canceled Previously") + ylab("Cancellation Rate") + labs(fill = "Has Canceled Previously")
```


We have seen that there is a stark difference in cancelation rates for guests with and without a previous cancelation on file, but is there any noticeable difference between the two hotels with regards to repeat cancelations?
```{r echo=FALSE}
previous_by_hotel = aggregate(is_canceled ~ has_previous_cancellation+hotel, data=hotels_bookings, mean)
vals <- c(previous_by_hotel[3])
combos <- c("No Prev. Cancellation, City Hotel", "Prev. Cancellation, City Hotel", "No Prev. Cancellation, Resort Hotel", "Prev. Cancellation, Resort Hotel")
d <- data.frame(combos, vals)
ggplot(d, aes(x=combos, y=is_canceled, fill=combos)) +  geom_bar(stat = "identity") + ggtitle("Plot Reservation Cancellation Rate by Hotel and Previous Cancellation") + xlab("Category") + ylab("Cancellation Rate") + labs(fill = "Category") +   theme(
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

We can see that just as it is higher overall, City Hotel's cancelation rate is higher than that of Resort Hotel regardless of whether or not the guest had previously canceled a reservation. Guests with track records of cancelation has no significant difference between the hotels, both should be aware of the high likelihood of those guests canceling again.

Another potential tool a hotel could use to anticipate whether a guest will follow through with their reservation is how far in advance the reservation was booked. If we break up every reservation into bins of less than a week in advance, between 1 week and 1 month, 1 and 2 months, 2 and 4 months, 4 and 6 months, 6 to 8 months, and 8+ months in advance, we get bins with at least 10,000 reservations in each category to look into which has the highest cancelation rate.  
```{r echo=FALSE}
hotels_bookings$lead_bin<-"240+ days" #default 
hotels_bookings$lead_bin[hotels_bookings$lead_time < 7 & hotels_bookings$lead_time >= 0] <- "<7 days"
hotels_bookings$lead_bin[hotels_bookings$lead_time < 30 & hotels_bookings$lead_time >= 7] <- "7-30 days"
hotels_bookings$lead_bin[hotels_bookings$lead_time < 60 & hotels_bookings$lead_time >= 30] <- "30-60 days"
hotels_bookings$lead_bin[hotels_bookings$lead_time < 120 & hotels_bookings$lead_time >= 60] <- "60-120 days"
hotels_bookings$lead_bin[hotels_bookings$lead_time < 180 & hotels_bookings$lead_time >= 120] <- "120-180 days"
hotels_bookings$lead_bin[hotels_bookings$lead_time < 240 & hotels_bookings$lead_time >= 180] <- "180-240 days"
hotels_bookings$lead_bin = factor(hotels_bookings$lead_bin, levels = c("<7 days", "7-30 days", "30-60 days","60-120 days","120-180 days","180-240 days", "240+ days"))

```

The table below indicates that the more lead time with which a reservation was booked, the more likely the party is to cancel it.
```{r echo=FALSE}
(lead_time_cancel <- aggregate(is_canceled ~ lead_bin, data = hotels_bookings, mean))
```

The graph below shows the cancelation rate at each time interval of booking. 

```{r echo=FALSE}
p <- ggplot(data=hotels_bookings, aes(x=lead_bin, y = is_canceled, fill=lead_bin)) + stat_summary(fun = "mean", geom="bar")
p + ggtitle("Plot Reservation Cancellation Rate by Lead Time") +
  xlab("Lead Time") + ylab("Cancellation Rate") + labs(fill = "Lead Time")

```

The cancelation rate crosses the mean cancellation threshold of 37% for the first time in the 60-120 days range, or between 2 and 4 months before the hotel stay date. It continues to grow from there, where the cancelation rate for a reservation more than 8 months out is around 64%, more than double the rate of reservations booked within a month. This trend is likely due to the fact that guests will make plans far in advance knowing that things may change between booking and the actual stay. If there isn't a penalty for cancelation or there is an insignificant one, guests will continue to book far in advance even if they do not fully intend to follow through on the reservation. 

#Conclusion


