---
title: 'TO414 Project 3: Spotify'
author: "Ankita Mohapatra"
date: "4/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Objective

This analysis is of interest to members of the Powerlifting and general strength sports community.

Powerlifting is a sport in which competitors compete to lift the most weight for their class in three separate barbell lifts: the Squat, Bench, and Deadlift. In this report, we will apply analytical methods to examine strength and yield some insights into the underpinnings of strength.

Intended Audience: Members of the powerlifting community: Federation Administrators, Coaches, Competitors, Strength Gym Owners, Powerlifting Gamblers, and Strength sport enthusiasts
Overarching Objective: Predict Strength
Underlying Objective 1 (for all audiences): Recognize which lifter and meet attributes are correlated to strength and success in meet lift attempts
Underlying Objective 2 (for competitors and their coaches): Select a federation that fits your goals, 
Data: Competitor data from powerlifting meets included in the OpenPowerlifting database as of April 2019. OpenPowerlifting is creating a public-domain archive of powerlifting history. 
Procedure: 

1. Descriptive Analytics: How does powerlifting performance vary across gender, age, and weight-based classes? Among the strongest lifters (adjusted for differences in gender, age, and weight), what are the key trends in equipment selection, drug testing, federation association, and meet event? Have lifters gotten stronger over the years? For regular competitors, what are the general strategies in weight jumps across attempts? Texas is a stronghold for Powerlifting - how do Texan lifters stack up against the overall data of lifters in terms of lifter profile and meet performance?
2. Predictive Analytics: Can we use the percent increase between attempts to predict whether the lifter will have a failed or succeeded attempt?  Can we look at trends behind weight jumps and estimate what percent increase the lifter will attempt from Attempt 2 to Attempt 3?**  Can we predict a lifter's best Squat or best Deadlift? 
3. Prescriptive Analytics: Should federations enforce drug testing? To maximize lift total without failing the next lift attempt, what is the optimal % jump in weight that a lifter should attempt?

Finally, we will identify several business management applications for the insights presented in this analysis.




# Data Preparation

## Pull & View Main Data
```{r}
spotify <- read.csv("data.csv")
#spotify_by_artist <- read.csv("data_by_artist.csv")
str(spotify)
#summary(spotify)

#audio_features <- read.csv("tracks.csv")
```
### Libraries
```{r}
packages <- c("dplyr", "class", "caret", "gmodels", "C50")
lapply(packages, require, character.only = TRUE)
```

### Data Manipulation
Main Data
```{r}
spotify <- subset(spotify, select = -c(id, name, release_date))
spotify$artists <- as.factor(spotify$artists)
spotify$key <- as.factor(spotify$key)
```

Spotify Main Set
```{r}
# Cut year into bins by decade
spotify$decades <- cut(spotify$year, breaks = c(0,1929, 1939, 1949, 1959, 1969, 1979, 1989, 1999, 2009, 2019, Inf), labels = c("20s", "30s", "40s", "50s", "60s", "70s", "80s", "90s", "2000s", "2010s", "2020s"))


# Classify popularity into bins by 75% and 90% percentiles
popularity_threshold <- 42
quantile(spotify$popularity, probs=seq(.1, .9, by = .1))
spotify$hit_class <- ifelse(spotify$popularity >= 56, "mega-hit", ifelse(spotify$popularity > popularity_threshold, "hit", "not hit"))
spotify$hit_class <- as.factor(spotify$hit_class)
# true or false if top 10% or not
spotify$binary_hit_class <- ifelse(spotify$hit_class == "mega-hit", 1, 0)
spotify$binary_hit_class <- as.factor(spotify$binary_hit_class)
table(spotify$hit_class)

```

Audio Features
```{r}
audio_features$release_year <- substr(audio_features$release_date,1,4)
audio_features$release_year <- as.numeric(audio_features$release_year)
audio_features$key <- as.factor(audio_features$key)
boxplot(audio_features$release_year, horizontal = T, main = "Year of Song Release")
audio_features <- subset(audio_features, select = -c(id, name, popularity, artists, id_artists, release_date))
```
```{r}
audio_features <- audio_features %>%
  select(danceability,everything())   #bring `danceability` to first column
```

## Normalized Sets
Normalize method
```{r}
normalize <- function(x){
  return ((x-min(x))/(max(x)-min(x)))
}
  
set.seed(1234)
```

### Spotify set

### Audio Features set
```{r}
audio_featuresMM <- as.data.frame(model.matrix(~.-1,audio_features))  #all columns are numeric
audio_featuresN<- lapply(audio_featuresMM,normalize) #apply normalize() to all columns
audio_featuresN <- as.data.frame(audio_featuresN)
```

## Test & Train
Un-normalized Train & Test Sets
```{r}
test_set <- sample(1:nrow(spotify), 30000)
spotify_train <- spotify[-test_set, ]
spotify_test <- spotify[test_set, ]

```

Normalized Train & Test Sets
```{r}
#audio_features
audio_train <- audio_featuresN[-test_set,]
audio_test <- audio_featuresN[test_set,]

#Response / Labels
audio_train_labs <- audio_featuresN[-test_set,"danceability"]
audio_test_labs <- audio_featuresN[test_set,"danceability"]
```


# General Music Attributes
## By Decade
```{r}
ggplot(spotify, aes(x = decades, y = tempo, fill=decades)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = decades, y = popularity, fill=decades)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = decades, y = valence, fill=decades)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = decades, y = danceability, fill=decades)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = decades, y = loudness, fill=decades)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = decades, y = explicit, fill=decades)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = decades, y = duration_ms, fill=decades)) + 
  geom_bar(stat = "summary", fun = "mean")
```
The graphs show us very interesting trends through the decades. There is a trend towards more loud, high tempo, exlicit, and dancible songs. It also shows that people on Spotify listen to music from the 90's and 2000's the most. 

## By Popularity
```{r}
hist(spotify$popularity)
```
```{r}
ggplot(spotify, aes(x = hit_class, y = danceability, fill=hit_class)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = hit_class, y = acousticness, fill=hit_class)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = hit_class, y = energy, fill=hit_class)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = hit_class, y = explicit, fill=hit_class)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = hit_class, y = instrumentalness, fill=hit_class)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = hit_class, y = liveness, fill=hit_class)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = hit_class, y = loudness, fill=hit_class)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = hit_class, y = mode, fill=hit_class)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = hit_class, y = speechiness, fill=hit_class)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = hit_class, y = tempo, fill=hit_class)) + 
  geom_bar(stat = "summary", fun = "mean")
```
```{r}
ggplot(spotify, aes(x = hit_class, y = valence, fill=hit_class)) + 
  geom_bar(stat = "summary", fun = "mean")
```

Based on the graphs above, we can see that mega-hits typically have higher dancibility, energy, explicitness, and tempo BUT lower acousticness, instrumentalness, liveness, loudness, mode, and valence.
```{r}
sort(table(spotify$artists), decreasing = TRUE)[1:20]
```


# Decade Prediction
## Is the song from the 80s?
```{r}
eighties_train <- subset(spotify_train, select = -c(artists, year, hit_class))
eighties_test <- subset(spotify_test, select = -c(artists, year, hit_class))
summary(eighties_train)
summary(eighties_test)
eighties_train$eighties <- ifelse(eighties_train$decades == "80s", 1, 0)
eighties_test$eighties <- ifelse(eighties_test$decades == "80s", 1, 0)
```

```{r}

eighties_log <- glm(eighties ~ acousticness + danceability + duration_ms + energy + explicit + 
                      instrumentalness + liveness + loudness + mode + 
                      speechiness + tempo + key + instrumentalness*danceability + 
                      duration_ms*acousticness + duration_ms*valence + duration_ms*valence + danceability*instrumentalness, 
                    data = eighties_train, family = "binomial")
summary(eighties_log)

eighties_pred_log <- predict(eighties_log, eighties_test, type = "response")
eighties_pred_log <- ifelse(eighties_pred_log > .23, 1, 0)
summary(eighties_pred_log)
confusionMatrix(as.factor(eighties_pred_log), as.factor(eighties_test$eighties))
```

```{r}

library(C50)

eighties_dtm <- C5.0(as.factor(eighties) ~ acousticness + danceability + duration_ms + energy + explicit + 
                      instrumentalness + liveness + loudness + mode + 
                      speechiness + tempo + key + instrumentalness*energy + 
                      duration_ms*acousticness + duration_ms*valence + duration_ms*valence + danceability*instrumentalness, data = eighties_train)


eighties_pred_dt <- predict(eighties_log, eighties_test, type = "response")
eighties_pred_dt <- ifelse(eighties_pred_dt > .5, 1, 0)
confusionMatrix(as.factor(eighties_pred_dt), as.factor(eighties_test$eighties))

```

## Which decades are easiest to predict?
```{r}
decade_dtm <- C5.0(decades ~ acousticness + danceability + duration_ms + energy + explicit + 
                      instrumentalness + liveness + loudness + 
                      speechiness + tempo + valence, data = spotify_train)

decades_pred <- predict(decade_dtm, spotify_test, type = "class")
summary(decades_pred)
confusionMatrix(decades_pred, as.factor(spotify_test$decades))
```


# Danceability Prediction
Spotify defines danceability as follows:
``Danceability describes how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity.
A value of 0.0 is least danceable and 1.0 is most danceable.''

(This was the first model we ran together)
```{r}
dance_model <- lm(danceability ~ tempo + valence + speechiness + 
                    mode + liveness + key + instrumentalness + 
                    explicit + energy + duration_ms + acousticness, data = spotify)
summary(dance_model)
```

By running regression models on the audio features dataset, we can try to predict to what extent the danceability algorithm incorporates the various elements such as those mentioned in the definition above.

```{r}
dm1 <- lm(danceability ~ ., data = audio_train)
```


### ANN

# Popularity Prediction
## What makes a song popular?
### Linear
```{r}
spotify_lin <- lm(popularity ~ valence + speechiness + 
                    mode + liveness + key + instrumentalness + 
                    explicit + energy + duration_ms + danceability + acousticness, data = spotify)
summary(spotify_lin)
```
### Log
```{r}
spotify_log <- glm(popularity_bin ~ valence + speechiness + 
                    liveness + instrumentalness + 
                    energy + duration_ms + danceability + acousticness,
                   data = spotify_train, family = "binomial")
summary(spotify_log)
spotify_pred <- predict(spotify_log, newdata = spotify_test, type = "response")
spotify_pred <- ifelse(spotify_pred >  popularity_threshold/100, 1, 0)
summary(spotify_pred)
CrossTable(x = spotify_pred, y = spotify_test$popularity_bin, prop.chisq=FALSE)
confusionMatrix(spotify_pred, spotify_test$popularity_bin)
```

### log model for hit class (mega-hit vs not) 
```{r}
log_test_labels <- spotify_test$binary_hit_class
log_test <- subset(spotify_test, select=-c(binary_hit_class))
log_train <- subset(spotify_train, select=-c(artists, popularity, year, hit_class))
log_model <- train(binary_hit_class ~ ., data = log_train, method = "glm", family="binomial")
summary(log_model)
```

#### Evaluate Log Model
```{r}
pred <- predict(log_model, log_test)
confusionMatrix(pred, log_test_labels)
```

### KNN
```{r}
knn_train <- subset(log_train, select=-c(decades, key))
knn_test <- subset(log_test, select=-c(artists, popularity, year, decades, hit_class, key))
knn_train <- as.data.frame(lapply(knn_train[1:(ncol(knn_train)-1)], normalize))
knn_test <- as.data.frame(lapply(knn_test[1:ncol(knn_test)], normalize))
knn_model_pred <- knn(train=knn_train, test=knn_test, cl=log_train$binary_hit_class, k=379)
```

#### Evaluate KNN Model
```{r}
confusionMatrix(knn_model_pred, log_test_labels)
```

## Holding artist constant, what makes a song popular?

# Conclusion
